Twenty years of dynamic occupancy modelling --- what’s been done, and where to next?

### 2024-04-12 Manuscript/domReview.qmd
Complete sections
completed manuscript sections rough text

@@

@@

A total of **\[number\]** articles were included in the review, **\[number\]** from the preliminary sample and **\[number\]** from the secondary sample. Based on the acceptance rate from the secondary sample, an estimated **\[hit number\]** of the **\[total number\]** articles citing @mackenzie2003 would have met inclusion criteria; i.e., fit a dynamic occupancy model.

![](Figures/CoveragePlot.jpeg){#fig-coverage}

## Study details

@@

Data for DOMs were collected from study areas on all continents and in **\[xx\]** countries, with the United States again representing a majority of the total articles published (@fig-StudyDetails a). These study areas ranged in size from less than 1 to over 1 million square kilometres

@@

## Data collection

## Modelling

@@

## Selection

@@

@@

**\<Covariate inclusion figure\>**

We provide guidelines for fitting dynamic occupancy models.

![](Figures/Guidelines.png){#fig-guidelines}

#

>>>>

## Article solicitation

@@

## Analysis

For each included article we collected data collected on authorship, stated research objectives, study taxa and field survey locations, data collection methods, covariates, and model selection and evaluation. Particular emphasis was placed on the covariates considered by authors, with no distinction between covariates which were or were not included in final models. As we were principally interested in the modelling process, our analysis did not directly explore final models or the results generated by those models. A spreadsheet of reviewed articles and collected data is provided in Appendix I.

@@

A total of **\[number\]** articles were included in the review, **\[number\]** from the preliminary sample and **\[number\]** from the secondary sample. Based on the acceptance rate from the secondary sample, an estimated **\[hit number\]** of the **\[total number\]** articles citing @mackenzie2003 would have met inclusion criteria. Total published articles have increased considerably over time.

![Estimated coverage of articles included in this review](Figures/CoveragePlot.jpeg){#fig-coverage}

@@

## Study locations, focal species, and data collection

Data for DOMs were collected from study areas on all continents and in **\[xx\]** countries, with the United States again representing a majority of the total articles published. These study areas ranged in size from less than 1 to over 1 million square kilometres (@fig-StudyDetails a).

Focal species were largely vertebrate taxa, although 6 articles fit models to invertebrates. A single non-animal taxa was represented in the sample, from an article fitting a DOM to a rare species of lichen @belinchón2017. [aa%] of articles modelled threatened taxa, and \[bb%\] modelled species invasive to the study area. The majority of articles (50, X%) fit single-species DOMs to one taxa. Of the remainder, 26 (X%) fit independent single-species models to more than one taxa. 8 articles fit (X%) multi-species implementations with explicit interactions between species (@fig-StudyDetails b).

Data used to fit dynamic occupancy models included human survey data such as transects or point counts (X% of articles), live-trapping data (X%), camera traps (X%) or bioacoustic monitors (Y%), and citizen science data (Z%); some articles incorporate multiple detection methods to build their detection datasets. Study duration, defined as the time elapsed between the first and last surveys conducted, ranged from 1 month to 40 years with the median study running for X years. The number of seasons ranged from 2 to Y with a median of Z; the number of observations per season ranged from X to Y with a median of Z. The number of sites surveyed per article ranged from X% to Y%, with the median study including Z sites (@fig-StudyDetails c).

@@

## Covariates

A summary of the covariates classes considered in the review sample is presented in @tbl-covariates. The most frequent

| Covariate type | Proportion of articles considering | Proportion including on Initial Occupancy | Proportion dynamic | Proportion directly observed | Frequency of interaction terms | Frequency of polynomial terms |
|-----------|-----------|-----------|-----------|-----------|-----------|-----------|
|                |                                    |                                           |                    |                              |                                |                               |
|                |                                    |                                           |                    |                              |                                |                               |
|                |                                    |                                           |                    |                              |                                |                               |
|                |                                    |                                           |                    |                              |                                |                               |
|                |                                    |                                           |                    |                              |                                |                               |
|                |                                    |                                           |                    |                              |                                |                               |
|                |                                    |                                           |                    |                              |                                |                               |
|                |                                    |                                           |                    |                              |                                |                               |
|                |                                    |                                           |                    |                              |                                |                               |
|                |                                    |                                           |                    |                              |                                |                               |

: Covariates considered in dynamic occupancy models {#tbl-covariates}

@@

@@

### Selection and evaluation

Bayesian models were far less likely to implement a model selection protocol compared to frequentist models: 30% of the former fit only a single model defined *a priori,* compared to 10% of the latter (@fig-selection). In frequentist models, stepwise selection protocols were the most popular choice. The vast majority of frequentist models (XX%) used AIC or AICc to identify the best model.

Objective evaluation of model fit was rare throughout the sample, regardless of selection protocol. XX% of articles calculated non-relative metrics, either via in-sample validation, bootstrapped goodness-of-fit tests, or out-of-sample validation.

@@

### Usage of dynamic occupancy models

Our review sample displays considerable diversity in implementations of dynamic occupancy models with widespread representation from mammals, birds, and herptiles. This assemblage contrasts with the related stochastic patch occupancy models (SPOMs, @gutiérrez-arellano), where the majority of models are fit to invertebrates, plants, and fungi. This may reflect SPOMs lack of consideration for imperfect detection and assumption of discrete habitat patches, limitations which are unlikely to be fulfilled by most vertebrate taxa.

DOMs also display great variation in scale, from local studies with a handful of sites to continental-scale studies spanning decades.

Several types of detection data have been successfully incorporated DOMs. This flexibility may be one of the more notable aspects of dynamic occupancy models; while the heirarchical data inputs may be more complex than those required by other models of species distributions, authors have been able to successfully manipulate datasets from many sources to meet requirements.

### Practices in the modelling process

Model selection approaches diverge between Bayesian and frequentist implementations of the model. While model selection will naturally look different between these two categories, it is unclear what effect a lack of model selection may have on the outputs produced by Bayesian models and it is likely that Bayesian model selection approaches should be further explored @hooten2015.

Notably, the popular sequential approach to model selection (where one sub-model is assessed at a time) may not provide optimal results as shown in @morin2020. Considerably more research is required on this topic to clarify how authors should best approach the model building process.

### Key considerations for model users

Implementation of dynamic occupancy models requires considerable thought to ensure that model outputs correspond with author expectations and ecological realities. Based upon our review of the existing literature we discuss key questions authors should carefully engage with during the modelling process.

1.  **Are seasons ecologically meaningful, and how do they affect the definition of occupancy?** As previously discussed, the duration of the season has important implications for both the reliability and proper interpretation of model results. Seasons which are not ecologically meaningful may run afoul of the closure assumption resulting in associated biases. Also relevant is how the definition of 'occupancy' may shift with season length and life-history of a species — for many species, occupancy within a year and within a week may be considerably different concepts, with shorter intervals more closely relating to the idea of 'use.' With the increasing proliferation of continuous detection methods such as camera traps and acoustic monitors where the definition of 'season' is to some degree arbitrary, it is increasingly important to thoughtfully consider how this element of the DOM is defined.

2.  **Are key drivers of occupancy and detection incorporated?** While it is effectively impossible to *fully* describe heterogeneity across a landscape, authors should make their best efforts to capture the drivers of occupancy and detection most likely to influence their study species. This is true even when another covariate is of principal interest to investigators — failure to include other major determinants of occupancy is likely to bias parameter estimates. One place where this is of particular importance is in determining drivers of detection, which is not only subject to observational conditions but also ecological determinants which may cause the species to spend more time at a location and thus be more frequently available for detection.

3.  **Ensure covariates reflect ecological complexity.** Many ecological relationships are non-linear and may require polynomials terms to be accurately represented. This is particularly true at large spatial scales where a larger portion of a species ecological niche is shown in surveyed data. Authors should question whether a linear term is sufficient to capture hypothetical relationships in their study systems and consider quadratic terms where appropriate.

4.   **Consider model evaluation.** Model evaluation remains rare in ecological research relative to other fields embracing complex modelling, likely due to a reluctance to 'throw away' data for use in testing. This concept should be challenged, particularly for larger scale long-term projects where reserving a fraction of data for testing is unlikely to substantially reduce precision of parameter estimates. For studies where this is genuinely not possible, reporting results of goodness-of-fit tests may still provide valuable information to readers.

### Priorities for future research

Work on dynamic occupancy modelling over the past 20 years has developed this model class into an important tool for describing species distributions and ecological relationships. Historically, methods-development research for dynamic occupancy models has focused on questions of model structure — namely ideas around the closure assumption and imperfect detection. This emphasis likely reflects the DOM's roots in the mark-recapture literature, where these topics have been discussed for several decades. Considerably less emphasis has been placed on the relation of model parameters to environmental factors via covariates and the model selection processes used to do so. This stands in stark contrast to related correlative species distribution models, where these topics have been the subject of considerable research over the years. There are opportunities for the DOM literature to learn from species distribution models, and more targeted research on the influence of the model selection process is necessary.

The use of DOMs in prediction has remained fairly limited despite significant potential for this use-case. Papers in our sample which have taken advantage of this potential included @mcgowan2020 who used DOMs to make predictions of black rail distributions under various management scenarios, and @saunders2020, who validated predictions from ecological niche models using continent-scale DOMs. Further implementations of these models may benefit from the aforementioned necessary research on model selection to ensure robust outputs.

A final opportunity for future research relates to autonomous detection methods, such as camera traps and bio-acoustic monitors — these technologies have permitted collection of enormous quantities of data particularly well-suited to dynamic occupancy models, but have also raised additional questions. The definition of seasons is of increased importance for these data types, where division of data into 'seasons' and 'observations' is arbitrary. Further research should explore the implications of this phenomenon and additional modelling opportunities made possible by this novel data type.

The past two decades have shown that dynamic occupancy models are an excellent tool for addressing a wide range of ecological questions relating to species occupancy throughout space and time. The flexibility afforded by model extensions and the considerable advantages of the class over other species distribution models is likely to ensure that these models continue to increase in popularity. Given their strong use on threatened taxa and excellent potential to support environemental management, it is important that research continues to support model users in creating the most reliable models possible.
